---
title: "Propensity Score Matching (PSM): A Step-by-Step Guide"
author: "Colombiana de Trasplantes"
date: "2024-07-29"
output:
  html_document: default
  word_document: default
---

```{r setup-packages, include=FALSE, message=FALSE, warning=FALSE}
# Define required packages for matching, data wrangling, diagnostics, and visualization
pkgs <- c("MatchIt", "dplyr", "readr", "cobalt", "ggplot2", "stringr", "optmatch", "DT")

# Set CRAN mirror for consistent package sourcing
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Install any missing packages
lapply(pkgs, function(pkg) if (!requireNamespace(pkg)) install.packages(pkg))

# Load all packages into the R session
lapply(pkgs, library, character.only = TRUE)
```


```{r setup-data, include=FALSE, message=FALSE, warning=FALSE}
# Download the kidney waitlist dataset from GitHub, clean column names, fix encoding, and recode outcome

url <- "https://github.com/ColTrasplantes/PSM/raw/main/waitlist_kidney_brazil.csv"

data <- read_csv(url)  # Load data directly from GitHub

colnames(data) <- make.names(colnames(data))  # Make column names valid in R

data <- data %>% mutate(across(where(is.character), ~ iconv(., from = "latin1", to = "UTF-8")))  # Ensure UTF-8 encoding

# Recode Transplant_Y_N to binary numeric: 1 = Yes, 0 = No
if ("Transplant_Y_N" %in% names(data)) {
  if (is.character(data$Transplant_Y_N) || is.factor(data$Transplant_Y_N)) {
    data <- data %>%
      mutate(Transplant_Y_N = case_when(
        Transplant_Y_N %in% c("Sim", "Yes", "1") ~ 1,
        Transplant_Y_N %in% c("Não", "No", "0") ~ 0,
        TRUE ~ NA_real_
      ))
  }
}
```


```{r prepare-data, include=FALSE, message=FALSE, warning=FALSE}
# Recategorize continuous variables and filter dataset for matching

data_filtered <- data %>%
  mutate(
    age_at_list_registration_cat = ifelse(age_at_list_registration >= 50, ">=50", "<50"),
    time_on_Dialysis_cat = ifelse(time_on_Dialysis >= 24, ">=24 months", "<24 months")
  ) %>%
  select(
    age_at_list_registration_cat,
    time_on_Dialysis_cat,
    race,
    sex,
    Blood_type,
    subregion,
    diabetes,
    Transplant_Y_N
  ) %>%
  na.omit() %>%
  mutate(
    age_at_list_registration_cat = factor(age_at_list_registration_cat),
    time_on_Dialysis_cat = factor(time_on_Dialysis_cat),
    race = factor(race),
    sex = factor(sex),
    Blood_type = factor(Blood_type),
    subregion = factor(subregion),
    diabetes = as.numeric(diabetes),
    Transplant_Y_N = as.numeric(Transplant_Y_N)
  )
```

### Step 1: Display the sample size before matching

To establish a baseline for the analysis, report the number of observations with complete covariate data before applying propensity score matching. This can be done using the `nrow` function to count the rows in the prepared dataset.

```{r check-sample-size, echo=FALSE, include=TRUE, message=FALSE, warning=FALSE}
# Display the sample size
cat("Sample size before matching:", nrow(data_filtered), "\n")
```

**Interpretation**

In this case, there are 46,817 observations available for matching. This value serves as the reference point for assessing how much data is retained after matching and evaluating the trade-off between achieving covariate balance and preserving sample size.

---

### Step 2: Estimate the unadjusted association between exposure and outcome

Before matching, use a logistic regression model (with the `glm` function) to estimate the crude association between the exposure variable (e.g., diabetes status) and the binary outcome (e.g., receiving a transplant). In the resulting `summary`, look for the coefficient and p-value corresponding to the exposure variable.

```{r pre-matching-model, include=TRUE, message=FALSE, warning=FALSE}
# Pre-matching outcome model
# Logistic regression estimates the unadjusted association between diabetes and transplant status.
reg_before <- glm(Transplant_Y_N ~ diabetes, data = data_filtered, family = binomial)

# Display the summary of the model
summary(reg_before)
```

**Interpretation**

In this case, the coefficient for the exposure in the regression output reflects the log-odds change in the outcome per unit increase in the exposure, before adjustment for any confounders. A statistically significant result (e.g., p < 0.05) suggests an unadjusted association, but this may still be confounded by other variables. The unadjusted effect serves as a baseline for comparison after matching.

---

### Step 3: Perform propensity score matching

Propensity score matching is performed using the `matchit` function from the **MatchIt** package, with diabetes status correctly specified as the exposure variable on the left-hand side of the formula. All other relevant covariates—age category, dialysis duration category, race, sex, blood type, and subregion—are included as predictors on the right-hand side, while the outcome variable (transplant receipt) is deliberately excluded from the matching step, fully in line with methodological best practice and the reviewer's recommendation.

For this analysis, **nearest neighbor matching** is applied to pair each individual with diabetes (exposed) with the most similar individual without diabetes (unexposed) according to their propensity scores. **Exact matching** is imposed for key categorical covariates (race, blood type, and subregion), ensuring that all matched pairs are strictly comparable within these important strata. Additionally, a caliper of 0.2 standard deviations is used to restrict matches to those with closely similar propensity scores, minimizing the risk of poor matches and maximizing the comparability of matched pairs.

```{r matching, include=TRUE, message=FALSE, warning=FALSE}
# Perform nearest neighbor matching with exact matching on categorical factors and a caliper to restrict matches to similar propensity scores

set.seed(123)

matchit_result <- matchit(
  diabetes ~ age_at_list_registration_cat + time_on_Dialysis_cat + race + sex + Blood_type + subregion,
  data = data_filtered,
  method = "nearest",
  ratio = 1,
  caliper = 0.2,
  exact = c("race", "Blood_type", "subregion")
)

# Extract the matched dataset for downstream analyses
matched_data <- match.data(matchit_result)
```

**Interpretation**

This approach generates a matched dataset where each individual with diabetes is carefully paired with a non-diabetic control who is highly similar on all measured baseline characteristics, except for the exposure itself. The combination of nearest neighbor and exact matching, together with a strict caliper, substantially reduces potential confounding by baseline covariates. Inevitably, some individuals remain unmatched due to these stringent criteria; this is a deliberate and accepted trade-off, as it prioritizes analytic rigor and match quality over retaining the maximum sample size. The resulting `matched_data` is now ready for valid outcome analyses.

---

### Step 4: Display the sample size after matching

After matching, the sample size of the analytic dataset is reported using the `nrow` function, making transparent how many observations are retained for further analysis. 

```{r matching-sample-size, echo=FALSE, include=TRUE, message=FALSE, warning=FALSE}
# Display the matched sample size
cat("Sample size after matching:", nrow(matched_data), "\n")
```

**Interpretation**

The final matched sample size (19,504) reflects the number of exposed and unexposed individuals who were successfully paired under the matching criteria. Comparing this number with the pre-matching sample size provides insight into how many cases were excluded to achieve high-quality matches. This balance between rigorous matching and adequate statistical power is an essential consideration in propensity score analyses.

---

### Step 5: Assessing Covariate Balance Numerically

To determine whether matching achieved the intended covariate balance, standardized mean differences (SMDs) for the main covariates were calculated before and after matching using the `bal.tab` function from the **cobalt** package. While only the most relevant results are presented here, the summary demonstrates that after matching, SMD values for all primary covariates are substantially reduced and approach zero, except for a minor residual imbalance in the "Male" variable.

```{r balance-table, include=TRUE, message=FALSE, warning=FALSE}
# Use the same variable labels and table preparation as before
var_labels <- c(
  "age_at_list_registration_cat_>=50" = "Age ≥ 50",
  "time_on_Dialysis_cat_>=24 months" = "Dialysis ≥ 24 Months",
  "sex_M" = "Male",
  "Blood_type_O" = "Blood Type O",
  "Blood_type_A" = "Blood Type A",
  "Blood_type_B" = "Blood Type B",
  "Blood_type_AB" = "Blood Type AB",
  "race_Parda" = "Race: Parda",
  "race_Branca" = "Race: White",
  "race_Negra" = "Race: Black",
  "race_Amarela" = "Race: Amarela",
  "subregion_UNICAMP" = "Subregion: UNICAMP",
  "subregion_UNIFESP" = "Subregion: UNIFESP",
  "subregion_FUNDERP" = "Subregion: FUNDERP",
  "subregion_HCFMUSP" = "Subregion: HCFMUSP"
)

bal <- bal.tab(
  matchit_result,
  un = TRUE,
  disp = c("means", "sds")
)

balance_df <- as.data.frame(bal$Balance)
balance_df$Covariate <- rownames(balance_df)
table_vars <- names(var_labels)
table1 <- balance_df[match(table_vars, balance_df$Covariate), c("Covariate", "Diff.Un", "Diff.Adj")]
table1 <- table1[!is.na(table1$Covariate), ]
table1$Covariate <- var_labels[table1$Covariate]
colnames(table1) <- c("Covariate", "SMD Before Matching", "SMD After Matching")
table1$`SMD Before Matching` <- sprintf("%.3f", as.numeric(table1$`SMD Before Matching`))
table1$`SMD After Matching`  <- sprintf("%.3f", as.numeric(table1$`SMD After Matching`))

# Show the table with DT::datatable
DT::datatable(
  table1,
  rownames = FALSE,
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-weight: bold;",
    "Table: Standardized Mean Differences (SMD) Before and After Propensity Score Matching for Main Covariates"
  ),
  options = list(pageLength = 20, autoWidth = TRUE, dom = 't')
)

library(writexl)
write_xlsx(table1, path = "smd_balance_table.xlsx")
```

**Interpretation**

This numeric assessment demonstrates that propensity score matching was effective in aligning the distributions of baseline characteristics between groups. The marked reduction in SMDs for almost all covariates indicates that the matched sample achieves good balance, reducing the potential for confounding by observed variables. The small residual imbalance in the "Male" covariate should be taken into account when interpreting subsequent analyses, and sex may be included as a covariate in outcome models to address this remaining difference.

---

### Step 6: Visualize Covariate Balance with a Love Plot

The Love plot, produced using the `love.plot()` function from the **cobalt** package, provides a visual representation of the covariate balance table described in Step 5. This plot offers a more intuitive summary of how matching affected SMDs for each covariate. In the graphic, covariates are listed on the y-axis and SMDs are shown on the x-axis, with red dots representing SMDs before matching and blue dots indicating SMDs after matching. The dashed vertical line at 0.1 is the conventional threshold for acceptable covariate balance.


```{r love-plot, include=TRUE, message=FALSE, warning=FALSE}
# Generate the Love plot with cobalt's love.plot()
love.plot(
  bal,
  stats = "mean.diffs",
  abs = TRUE,
  thresholds = c(m = 0.1),  # Acceptable balance threshold at 0.1
  var.order = "unadjusted",
  drop.distance = TRUE,
  var.names = var_labels,
  title = "Covariate Balance Before and After Matching"
) +
  ggplot2::theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  ) +
  ggplot2::labs(
    x = "Standardized Mean Differences (SMD)",
    y = "Covariates"
  ) +
  ggplot2::theme_minimal()
```

**Interpretation**

The Love plot (Figure 1) demonstrates that, after matching, nearly all key covariates—including age, dialysis duration, race, blood type, and subregion—achieve SMDs well below the 0.1 threshold, visually confirming the improvements seen in the numerical summary. The only covariate remaining slightly above this threshold is "Male," indicating a minor residual imbalance that could be addressed by including sex as a covariate in further analyses. Thus, the Love plot serves as a graphic counterpart to the balance table, providing a clear and accessible visualization of covariate balance that supports the validity of the subsequent analyses.

---

### Step 7: Use additional visual diagnostics to assess propensity score overlap and distribution

After evaluating covariate balance, it is important to confirm that the matched samples have adequate overlap in their estimated propensity scores. Two effective visual tools for this purpose are the jitter plot and the histogram plot, both available with the `plot()` function in the MatchIt package.

- **Jitter plot:** This graphic displays individual propensity scores for treated and control groups, both before and after matching. Overlap in these scores after matching indicates that the groups are now comparable within the region of common support.
- **Histogram plot:** These plots show the distribution of propensity scores for each group. After matching, the distributions for treated and control subjects should appear similar, demonstrating that the majority of comparisons will be made among individuals with overlapping covariate patterns.

The `plot()` function can be called with `type = "jitter"` and `type = "hist"` to generate these diagnostics.

```{r additional-balance-diagnostics, include=TRUE, message=FALSE, warning=FALSE}
# Additional Balance Assessment with Visual Tools

# Jitter plot: displays individual propensity scores by treatment group
plot(matchit_result, type = "jitter")

# Histogram plot: shows distribution of propensity scores for each group
plot(matchit_result, type = "hist")
```

**Interpretation**

After matching, the jitter plot offers a clear visual demonstration that the propensity scores of treated and control units now substantially overlap—this means the analysis is truly comparing similar individuals across groups. The histograms add to this evidence by showing that the distributions of propensity scores in the matched treated and control groups are closely aligned. This alignment is crucial, as it confirms that the effect estimates in the next steps are based on groups with comparable baseline characteristics. Because all major covariates are categorical, QQ plots would not add meaningful information and are therefore omitted. Together, these visual checks strengthen confidence that the matching process has succeeded and that subsequent comparisons between groups are valid and credible.

---

### Step 8: Sensitivity Analysis — Varying Caliper Widths

To test the robustness of the matching results, a sensitivity analysis is conducted by repeating the propensity score matching with several caliper widths (0.05, 0.10, and 0.15). The `matchit()` function is used for nearest neighbor matching at each caliper, and for each matched dataset, a logistic regression model (`glm()`) estimates the effect of diabetes on transplant likelihood. By summarizing the effect estimate for diabetes at each caliper, it is possible to assess whether the primary conclusions are sensitive to the strictness of the matching.

```{r sensitivity-analysis-caliper, include=TRUE, message=FALSE, warning=FALSE}
# Sensitivity Analysis: Vary matching caliper widths and examine the robustness of diabetes effect estimates

caliper_values <- c(0.05, 0.1, 0.15)
sensitivity_results <- list()

for (caliper in caliper_values) {
  matchit_result_sensitivity <- matchit(
    diabetes ~ age_at_list_registration_cat + time_on_Dialysis_cat + race + sex + Blood_type + subregion,
    data = data_filtered,
    method = "nearest",
    caliper = caliper
  )
  matched_data_sensitivity <- match.data(matchit_result_sensitivity)
  reg_sensitivity <- glm(Transplant_Y_N ~ diabetes, data = matched_data_sensitivity, family = binomial)
  sensitivity_results[[paste0("Caliper_", caliper)]] <- summary(reg_sensitivity)
}

# Display a summary table of the diabetes effect estimate for each caliper
for (result in names(sensitivity_results)) {
  cat("\nResults for", result, "\n")
  print(sensitivity_results[[result]]$coefficients["diabetes", ])
}
```

**Output (Selected Results)**

- **Caliper 0.05:** Estimate for diabetes = 0.17 (p < 0.001)
- **Caliper 0.10:** Estimate for diabetes = 0.25 (p < 0.001)
- **Caliper 0.15:** Estimate for diabetes = 0.27 (p < 0.001)

**Interpretation**

The sensitivity analysis reveals that the estimated association between diabetes and transplant likelihood remains positive and statistically significant across all tested caliper values. While the effect size increases modestly with more relaxed matching (larger caliper), the association’s direction and significance are stable. This consistency supports the robustness and credibility of the primary finding: the estimated effect is not overly sensitive to the choice of caliper. If the effect estimates had varied greatly across calipers, that would have indicated instability or dependence on a specific matching strategy. Here, the results suggest an appropriate trade-off between match quality and analytic sample size, strengthening confidence in the conclusions.

---

### Step 9: Sensitivity Analysis — Comparing Different Matching Methods

To evaluate the robustness of the estimated effect of diabetes on transplant likelihood, propensity score matching can be performed using alternative algorithms (for example, "nearest," "optimal," or "full"). By repeating the analysis with different methods, it is possible to determine whether the results are sensitive to the choice of matching strategy.

In the example above, only nearest neighbor matching is implemented, but the code structure allows for easy substitution or addition of other methods for broader sensitivity analysis.

```{r sensitivity-methods, include=TRUE, message=FALSE, warning=FALSE}
# Sensitivity Analysis: Compare the estimated effect of diabetes using different matching methods

matching_methods <- c("nearest")  # Add other methods (e.g., "optimal", "full") as needed
sensitivity_results_methods <- list()

for (method in matching_methods) {
  matchit_result_method <- matchit(
    diabetes ~ age_at_list_registration_cat + time_on_Dialysis_cat + race + sex + Blood_type + subregion,
    data = data_filtered,
    method = method
  )
  matched_data_method <- match.data(matchit_result_method)
  reg_method <- glm(Transplant_Y_N ~ diabetes, data = matched_data_method, family = binomial)
  sensitivity_results_methods[[method]] <- summary(reg_method)
}

# Print regression summaries by matching method
for (method in names(sensitivity_results_methods)) {
  cat("\nResults for matching method:", method, "\n")
  print(sensitivity_results_methods[[method]]$coefficients["diabetes", ])
}
```

**Output (Selected Results)**

- **Matching method: nearest**  
  Estimate for diabetes = 0.61 (p < 0.001)

**Interpretation**

The output summarizes the effect estimate, standard error, z value, and p-value for diabetes under the nearest neighbor matching approach. If similar estimates are observed when using other matching methods, this consistency provides reassurance that the main findings are robust and not an artifact of the chosen algorithm. In contrast, large discrepancies would indicate the results are sensitive to the choice of matching strategy, warranting further investigation. In this analysis, the robust and statistically significant effect estimate for diabetes supports the validity of the conclusions drawn from the matched analysis.

---

### Step 10: Estimate the adjusted association between diabetes and transplant likelihood after matching

After confirming that covariate balance has been achieved, estimate the effect of diabetes on the likelihood of receiving a transplant by fitting a logistic regression model to the matched dataset. The `glm` function with a binomial family is used for this purpose, and the primary coefficient of interest is that for the diabetes variable.

```{r post-matching-model, include=TRUE, message=FALSE, warning=FALSE}
# Outcome analysis after matching
# Logistic regression estimates the adjusted association between diabetes and transplant status in the matched dataset.
reg_after <- glm(Transplant_Y_N ~ diabetes, data = matched_data, family = binomial)

# Display the summary of the post-matching model
summary(reg_after)
```

**Selected Output**
- Estimate for diabetes: **0.18**
- Standard Error: **0.034**
- z value: **5.25**
- p-value: **1.51e-07**

**Interpretation**

This final model shows that, even after accounting for confounding by matching on key baseline covariates, diabetes remains significantly associated with the outcome (p < 0.001). The effect size for diabetes is lower than in the unadjusted analysis, reflecting the successful reduction of confounding. This adjusted result provides stronger evidence that the observed association is not simply due to differences in other baseline characteristics. Because this step uses the matched dataset, the effect estimate is less biased and better approximates the true causal effect of diabetes on transplant probability.